\documentclass[12pt]{memoir}

\usepackage{cdsc-memoir}
% there are two chapter styles: cdsc-article and cdsc-memo
% memo assumes that you remove the "\\" and the email address from the
% \author field below as well as that you will comment out the
% \published tag
\chapterstyle{cdsc-article}

\usepackage[utf8]{inputenc}
\usepackage{wrapfig}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage[garamond]{mathdesign}


% \textwidth = 5.2inches
\usepackage[letterpaper,left=1.65in,right=1.65in,top=1.3in,bottom=1.2in]{geometry}

% packages i use in essentially every document
\usepackage{graphicx}
\usepackage{enumerate}

% packages i use in many documents but leave off by default
% \usepackage{amsmath, amsthm, amssymb}
% \usepackage{dcolumn}
% \usepackage{endfloat}

% import and customize urls
\usepackage[usenames,dvipsnames]{color}
\usepackage[table]{xcolor}
\usepackage{soul}
\usepackage[breaklinks]{hyperref}
\usepackage{array}
\hypersetup{colorlinks=true, linkcolor=Black, citecolor=Black, filecolor=Blue,
    urlcolor=Blue, unicode=true}
\hyphenation{com-men-sal-ism}
% list of footnote symbols for \thanks{}
\makeatletter
\renewcommand*{\@fnsymbol}[1]{\ensuremath{\ifcase#1\or *\or \dagger\or \ddagger\or
 \mathsection\or \mathparagraph\or \|\or **\or \dagger\dagger
  \or \ddagger\ddagger \else\@ctrerr\fi}}
\makeatother
\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}

\DeclareRobustCommand{\myhl}[2]{\sethlcolor{#1} \hl{#2}}

% add bibliographic stuff 
\usepackage[american]{babel}
\usepackage{csquotes}
\usepackage[natbib=true, style=apa, backend=biber]{biblatex}
\addbibresource{Ecology.bib}
\DeclareLanguageMapping{american}{american-apa}

\defbibheading{secbib}[\bibname]{%
  \section*{#1}%
  \markboth{#1}{#1}%
  \baselineskip 14.2pt%
  \prebibhook}

\def\citepos#1{\citeauthor{#1}'s (\citeyear{#1})}
\def\citespos#1{\citeauthor{#1}' (\citeyear{#1})}

% memoir function to take out of the space out of the whitespace lists
\firmlists

% LATEX NOTE: these lines will import vc stuff after running `make vc` which
% will add version control information to the bottom of each page. This can be
% useful for keeping track of which version of a document somebody has:
% \input{vc}
% \pagestyle{cdsc-page-git}

% LATEX NOTE: this alternative line will just input a timestamp at the
% build process, useful for Overleaf
% \pagestyle{cdsc-page-overleaf}
\usepackage{tikz}
\usetikzlibrary{positioning,shapes,arrows,fit,chains,shapes.geometric,trees}

\definecolor{hypothesizedcell}{HTML}{a9d89c}
\definecolor{nullhypothesizedcell}{HTML}{d89ca9}
\definecolor{agnosticcell}{HTML}{9ca9d8}

\begin{document}
\renewcommand{\arraystretch}{2}
\setlength{\parskip}{4.5pt}
% LATEX NOTE: Ideal linespacing is usually said to be between 120-140% the
% typeface size. So, for 12pt (default in this document, we're looking for
% somewhere between a 14.4-17.4pt \baselineskip.  Single; 1.5 lines; and Double
% in MSWord are equivalent to ~117%, 175%, and 233%.

\baselineskip 16pt

\title{Outline for: Unpacking Ecological Approaches to Online Communities }
\author{Your Name\\
        \href{mailto:youremail@uw.edu}{youremail@uw.edu}}
\date{}

\published{\textsc{\textcolor{BrickRed}{This document is an
  unpublished draft.\\ Please do not distribute or cite without
  permission.}}}

\maketitle

\begin{abstract}
Internet researchers are increasingly aware of important interdependence between online groups.  HCI researchers have drawn density dependence theory from population ecology to theorize competitive and complementary dynamics between communities with overlapping content and participants and predict growth and survival of online groups.   Density dependence theory highlights similarities between online groups as exchangeable members of a homogeneous population while minimizing the role of diversity and the importance of specialization.   Community ecology, a complementary approach proposes that each group plays a distinctive role in the broader sociotechnical ecosystem.   Vector autoregression (VAR) models are a method from community ecology  for estimating a network of competitive, complementary, and asymmetrical commensalist relations.  We introduce community ecology to HCI research by comparing population and community ecologies, explicating applicable concepts from community ecology, and testing hypothesis using a vector autoregression model fit to time-series data from Reddit.
\end{abstract}

\section{Dissemination Objectives}

This is a dissertation chapter and a CHI 2021 paper. The CHI 2020 deadline is Sep. 10 2020 for abstracts and metadata and Sep. 17th for the submission. I aim to submit or have a solid draft before then. Let's say July 1. 

I'll release code and datasets.

There isn't a practitioner or outreach component specific to this paper. 

\section{Rationale}

\begin{itemize}
\item Accumulating empirical results demonstrate how interdependence between online groups shapes participation, growth, and survival (we use the term ``online group'' to avoid confusion with ecological communities).  Yet relatively few studies provide general methodological tools and even fewer provide general theoretical concepts for systematically studying interdependence between online groups. 

\item Ecology provides general tools for conceptualizing and modeling these relationships.  Organizational social scientists have long worked on extensions from biological ecology and influenced prior research in HCI.  We continue building ecological theorizing in the context of HCI by introducing concepts and models from community ecology.  Community ecology aims to characterize the network of relations that shift or stabilize an ecosystem and thereby gain insights into the nature of the system and how it can be regulated.  

\item Prior work in HCI tests theories of density dependence from population ecology based on overlapping users and content \citep{zhu_impact_2014,zhu_selecting_2014,wang_impact_2013}.  Population ecology places online groups in the position of the biological organism and the set of groups on an online platform occupy the position of a population of species.  This highlights the similarity of groups as members of the same populations while minimizing the diverse roles that groups can play in the online ecosystem.   Community ecology by contrast is a more holistic ecological approach that analyzes diverse populations as interdependent constituents of an ``ecological community'' \citep{astley_two_1985-1}.  Community ecology places each community in the conceptual position of a distinct biological population and seeks to understand how communities relate to one another.  

In their empirical analyses, the above studies already take steps toward a community ecology approach by modeling density dependence as structured by overlapping memberships and topics between online groups.  Rather than assuming all communities face the same environment in strict population ecology approach, each group in their analysis faces a distinct environment of overlapping groups. They use the ``niche'' concept to refer to fit between a group and it's participants and content. Yet they stop short of explicating community ecology concepts or estimating the ``community matrix'' which quantifies relations of interdependence between online communities. 

% Why are community ecology concepts useful to understanding online communities?
\item Community ecology provides concepts for how online groups relate to one another.  These can give community managers ways to think about the role of individual communities in the broader context, to help answer questions like ``How might groups about similar topics shape activity and participation in one another?'' or ``What topics might we be able to create a new group around?'', ``When will we have more small groups or few large groups?'' Community ecology can also inform design choices like whether to allow users to create new groups and the potential effects of interventions like banning, quarantining, and norms against coordinated cross-participation.

% Why do we care about estimating the community matrix?
% Why is estimating the CM a goal of community ecology?
\item Community ecology provides statistical models tools for understanding sociotechnical ecosystems.  In particular, Vector autoregression (VAR) models can infer ecological networks of complementary, competitive, predatory, and communal relationships.  VAR analysis can quantify the stability of the system and affords exploration of counterfactual forecasts to simulate hypothetical interventions. 

\item Geographically local online communities are important sites in the networked public sphere where people go to make sense of current events and local concerns.  They also generate data that is relatively easy to cast into a community ecology model.  This study contributes to understanding these communities through the lens of community ecology. 

\item Community ecology promises a vast depth of analysis at levels of granularity ranging from narrow groups of a handful of communities to estimating ecological networks for entire large platforms like Reddit.  Therefore this project will be a first step focused on translating community ecology for HCI and a demonstrative analysis.  

\end{itemize}

\section{General Objectives}

\begin{itemize}
\item Provide answers to ``How does the growth/success of an online group depend on the environmental context?''  and ``How can we understand interdependence between online groups?''

\item Provide an overarching conceptual framework for studies of interdependence in social computing.

\end{itemize}

\section{Specific Objectives}
\begin{itemize}

\item Argue that community ecology is a more appropriate framework than population ecology for thinking about interdependence between online groups.

\item Build on prior use of the concept of \emph{niches} in HCI to emphasize that niche overlaps can be associated qualitatively different kinds of relationships.   

\item Define and motivate concepts of \emph{competition}, \emph{complementary} and \emph{comensalism} and \emph{specialization} for online communities.

\item Explicate the above concepts through testing hypotheses about when communities will compete or be complementary using \emph{resource partitioning theory} and \emph{the principle of competitive exclusion} using one or more case studies from Reddit.

\end{itemize}

\section{Propositions}

\begin{itemize}
\item \textbf{P1:} For geographic subreddits, ecological relationships between generalists will probably be competitive. That is, generalist $\rightarrow$ generalist dyads will have positive coefficients in estimated community matrices. 

\item \textbf{P2:} For geographic subreddits, ecological relationships from generalist to specialists will probably be beneficial to specialists. That is, generalist $\rightarrow$ specialist dyads will have positive coefficients in estimated community matrices.


\item \textbf{P3:} For geographic subreddits, ecological relationships between generalists and specialists are likely to be commensal. Sepcifically, generalist $\rightarrow$ specialist dyads will have greater coefficients in estimated community matrices compared to specialist $\rightarrow$ generalist dyads. 

\end{itemize}

\section{Hypotheses}
\begin{itemize}

% Competitive exclusion / specialization means that online  
\item \textbf{H1:} Changes in the number of user accounts posting in a generalist group will be negatively associated related with changes in the number of user accounts posting in a generalist group  \footnote{``Association'' in these hypotheses refers to marginal correlation, i.e. signs in the community matrix}.

\item \textbf{H2:} Changes in the number of user accounts posting in a generalist group will be more positively associated with changes in the number of user accounts posting in a specialist group than visa-versa.

\item \textbf{H3:} Changes in the number of user accounts posting in a generalist group will be more strongly associated with changes in the number of user accounts posting in a specialist group than visa-versa.

\end{itemize}

\section{Conceptual Model}

\label{sec:conc-modeld}

The argument for \textbf{P1} and \textbf{H1} introduces competitive dynamics.  Two different generalist subreddits can exist about the same topical area, but they are likely to compete with one another.  By the principle of competitive exclusion the competition must be weak enough that one group doesn't collapse and therefore we should expect for some elements of differentiation.


\begin{table}[h]
\input{resources/H1_table.tex}
\end{table}

In the case of /r/seattle and /r/seattlewa evidence of differentiation includes the custom look-and-feel, and different and statistically significant differences in the distribution of votes and comments across posts. /r/seattle has more votes and less comments per post compared to /r/seattlewa suggesting that it is more focused on social content ranking and less on discussion. Yet topically the two groups are both broadly scoped generalists and therefore we anticipate that a bi-directional competitive relationship is most likely.  We also expect competition between these two groups based on the historical origin of /r/seattlewa, for it was created immediately in response to conflict between the moderator of /r/seattle and community members.

\input{resources/H1_tikz.tex}

A mutualistic relationship between the two subreddits is counter-intuitive because participants in both communities must split their efforts and attention across both. Mutualism might be possible because the communities also share content, but \textbf{H1} proposes that competition over attention is more relevant. 

Relationships analogous to predation are also plausible. This might happen if participants often move between communities through a pipeline such that they commonly discover /r/seattle when they first go looking for a Seattle-based online group, but soon learn that /r/seattlewa is really the more active discussion group.   Then increases in participation in /r/seattle will be associated with subsequent increases in participation in /r/seattlewa and increased participation in /r/seattlewa  is associated with decreases in participation in /r/seattle.

A competitive commensalist relationship is also plausible if increases in one community are associated with decreases in the other, but increases in the second community have no association with changes in the first.  This might happen  /r/seattlewa attracts participants who would have participated in /r/seattle if not for /r/seattlewa, but the converse isn't true and participants in /r/seattlewa will only contribute to /r/seattlewa whether /r/seattle exists or not.  In this scenario an increase in /r/seattlewa will be associated with a future decrease in /r/seattle, but an increase in /r/seattle will be statistically independent of future changes in /r/seattlewa.

A second way that a commensal competitive relationship can happen is if bidirectional competition is mixed with commensal mutualism. This could happen if /r/seattle and /r/seattlewa compete over participants, but activity in /r/seattle somehow generates activity in /r/seattlewa. It's hard to see how this might happen in this particular example, but Reddit's so big it probably happens somewhere. 

\input{resources/H2_table.tex}

% Diagram for \textbf{H1} goes here
The argument for \textbf{P2} and \textbf{H2} is based on the theories of resource partitioning and the principle of competitive exclusion.  Specialization is a good strategy because specialists can avoid competition with generalists.  In the specific case of local geographic subreddits, we expect that generalist groups (e.g. /r/seattle) will attract and sustain participants who are also cyclists.  If the cyclists form their own specialized group (e.g. /r/seattlebikes), will this community compete with the generalist community?  According to resource partitioning theory which posits that specialists will survive when they have competitive advantages over generalists.  This leads to \emph{resource partitioning} as content or discussion of /r/seattlebikes should have less overlap with that of /r/seattle.  

%If this competition is strong that will not be a good sign for the survival of /r/seattlebikes   For example, we expect that if /r/seattlebikes can sustain an active community that means that /r/seattlebikes participants are able to have interactions, have conversations, or find content related to their interests in ways they prefer to /r/seattle.  So while it is possible for a specialist to compete with a generalist it is unlikely that this competition hurts the specialist very much.

\input{resources/H2_tikz.tex}
Specialists are unlikely to survive for a second reason. If a specialist occupies a niche so narrow that it is difficult to maintain a critical mass of participants then it will fail to sustain activity.  Helpful generalists can create a hospitable environment for specialists by routing them participants.  This can be seen in practice by cross linking. On the other hand it seems more likely that increases in activity in the generalist will help the specialist.  Increasing activity in /r/seattle might be associated with growing participation on Reddit in the Seattle area in general.  Users of both communities might link to /r/seattlebikes in bike-related threads in /r/seattle and growth in /r/seattle increases the size of the audience that would be exposed to these messages.


% Diagram for \textbf{H2} goes here

What about the effect of specialist subreddits on generalist ones?  It seems hard to imagine that an increase in activity on /r/seattlebikes will cause a large decrease in participation in /r/seattle because discussions about bikes in /r/seattle constitute a relatively small proportion of activity.  Commensal relations are common in biology and we expect them to be common in the realm of online groups as well.


\textbf{H2} predicts how a specialist depends on a generalist, but says nothing about how the generalist's course is shaped by the specialist.    \textbf{H3} says that the effect of specialists on generalists is weak relative to the effect of generalists on specialists.  This is based on the intuition that while generalists' niches generally cover specialists' niches specialists' niches cover only a small part of generalists' niches.  \textbf{H3} is agnostic as to the sign of the relationship, but taken with \textbf{H2} it's clear that we hypothesize mutualistic commensalism between generalists and specialists.

\input{resources/H3_table.tex}

% \begin{table}[t]
% \small
%   \begin{tabular}{*{4}{|m{0.25\textwidth}}|}
%     \hline
%     & generalist hurts \newline specialist & generalist helps \newline specialist & generalist has no \newline effect on specialist \\ \hline
%     specialist hurts \newline generalist & \cellcolor{nullhypothesizedcell} competition \newline increased chance of inactive specialist & \cellcolor{hypothesizedcell} Specialist ``preys'' on generalist, siphoning off participants & \cellcolor{agnosticcell} Competitive \newline commensalism, \newline surprise!  \\
%     specialist helps \newline  generalist & \cellcolor{nullhypothesizedcell} Generalist ``preys'' on specialist, \newline siphoning off participants  & \cellcolor{hypothesizedcell} Mutualism & \cellcolor{agnosticcell} Mutualistic \newline commensalism \\
%     specialist has no \newline effect on generalist & \cellcolor{nullhypothesizedcell} Competitive \newline commensalism &  \cellcolor{hypothesizedcell} Mutualistic \newline commensalism & \cellcolor{agnosticcell} No relationship \\ \hline
% \end{tabular}
% \end{table}



%Organizational ecologists \citet{hannan_logics_2007} argue that ecological dynamics between organizations depend on the ``audiences'' of organizations by which they mean groups like investors, employees, and consumers.  In applying this thinking to generalization and specialization of online communities, we proposes that for a specialist community to survive it must have an audience of participants who prefer to participate in the specialist community over the other.  



For \textbf{}


\clearpage



% bibliography here
\setcounter{biburlnumpenalty}{9001}
\printbibliography[title = {References}, heading=secbib]


\end{document}

%  LocalWords:  
